Document S1

### 2025-09-03 latex/docS1.tex

@@

To analyse how well the ARG reflects the phylogenetic structure implicit in the Pango lineage naming system,
we considered the Pango assignments generated by Pangolin on the alignments for each node in the ARG (STAR Methods).
If the ARG perfectly reflected the Pango lineage structure,
each lineage would form a clade descending from a single originating node with that label.
We identified putative origination nodes among those labeled with a given lineage as the node
with the maximum number of descendant samples and the earliest time (if there is a tie).

@@

\subsection*{Analysis of the Jackson et al. (2021) recombinants}
Jackson et al. (2021) conducted detailed analyses on some early recombinants
involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
The authors first performed a targeted search for candidate recombinants by
scanning sampled sequences for sequence motifs associated with Alpha, followed
by an analysis to reduce the list of plausible recombinants displaying signs of
onward transmission. Here we compared our sc2ts detection results of these
recombinants with the detection results reported by Jackson et al.

% Data derived from the notebooks/jackson_recombinants.ipynb notebook
Jackson et al.\ described four groups of recombinants (named A to D) and
four singleton recombinants.
Of the 16 samples analyzed by Jackson et al, 12 were included in the ARG;
the remaining 4 were excluded because the sequences did not pass the 
QC filtering criteria, or due to a high HMM cost. Groups A-D are summarized 
in Table~\ref{tab:jackson} and show excellent concordance in terms 
of the parental Pango lineages and breakpoint intervals. All four groups
are directly associated with strongly supported recombination events 
in the ARG in terms of the number of averted mutations 
(A: 16, B: 6, C: 30, D: 18).
The singleton CAMC-CBA018 has 
the parents B.1.177+B.1.1.7 and the breakpoint interval 16,177--20,133 
in the ARG 
and is strongly supported by 14 averted mutations.
The Jackson et al.\ results are almost identical with the same parent lineages, 
and breakpoint interval of interval 11,395–21,993.

the parents B.1.177.16+B.1.1.7 and a breakpoint interval of 24915--27972 
in the ARG, again strongly supported by 11 averted mutations.
The Jackson et al.\ results for this lineage has the same parents,
with the breakpoint interval 26,800–-27,974.
The singleton QEUH-1067DEF has the parents B.1.1.7+B.1.177.9 
and a breakpoint interval of 7,729--10,870 in the ARG, 
supported by 11 averted mutations.
The Jackson et al.\ results for this lineage has the same parents,
with the breakpoint interval 6,953--10,872.

>>>>

\subsection{Major deletions in the ARG}
% Referenced by the main Results.
Basically, we want to say we did a lot of work to integrate signals of major indels into the ARG,
making it an even more powerful resource overall.
Also, indels are encoded by the Mutation-Annotated Tree format used by UShER phylogenies (though PanMAT does this).

% See https://github.com/jeromekelleher/sc2ts-paper/issues/472
At minimum, we want to show compare the frequencies of the top N indels in the ARG and those reported in Li et al. (2023).
This can be shown in a table.

@@

To analyze how well the ARG reflects the phylogenetic structure implicit in the Pango lineage naming system,
we considered the Pango assignments generated by Pangolin on the alignments for each node in the ARG (STAR Methods).
If the ARG perfectly reflected the Pango lineage structure,
each lineage would form a clade descending from a single originating node with that label.
We identified putative origination nodes among those labeled with a given lineage as the node
with the maximum number of descendant samples and the earliest time (if there is a tie).

@@

\subsection*{Analysis of the Jackson et al. recombinants}
Jackson et al.\cite{Jackson2021} conducted detailed analyses of some early recombinants
involving B.1.1.7 (Alpha), including the first designated Pango X lineage, XA.
The authors first performed a targeted search for candidate recombinants by
scanning sampled sequences for sequence motifs associated with Alpha, followed
by an analysis to reduce the list of plausible recombinants displaying signs of
onward transmission. Here we compared our sc2ts detection results of these
recombinants with the detection results reported by Jackson et al.

% Data derived from the notebooks/jackson_recombinants.ipynb notebook
% See https://github.com/jeromekelleher/sc2ts-paper/issues/293
Jackson et al. described four groups of recombinants (named A to D) and
four singleton recombinants.
Of the 16 samples analyzed by Jackson et al, 12 were included in the ARG;
the remaining 4 were excluded because the sequences did not pass the 
QC filtering criteria, or due to a high HMM cost. Groups A-D are summarized 
in Table~\ref{tab:jackson} and show excellent concordance in terms 
of the parental Pango lineages and breakpoint intervals. All four groups
are directly associated with strongly supported recombination events 
in the ARG in terms of the number of averted mutations 
(A: 16, B: 6, C: 30, D: 18).
The singleton CAMC-CBA018 has 
the parents B.1.177+B.1.1.7 and the breakpoint interval 16,177--20,133 in the ARG, 
and is strongly supported by 14 averted mutations.
The Jackson et al.\ results are almost identical with the same parent lineages, 
and breakpoint interval of interval 11,395–21,993.
The singleton MILK-103C712 has 
the parents B.1.177.16+B.1.1.7 and a breakpoint interval of 24915--27972 
in the ARG, again strongly supported by 11 averted mutations.
The Jackson et al.\ results for this lineage has the same parents,
with the breakpoint interval 26,800–-27,974.
The singleton QEUH-1067DEF has the parents B.1.1.7+B.1.177.9 
and a breakpoint interval of 7,729--10,870 in the ARG, 
supported by 11 averted mutations.
The Jackson et al.\ results for this lineage has the same parents,
with the breakpoint interval 6,953--10,872.
